First, I would like to thank Sandra K\"ubler. I could not have asked
for a better, more supportive advisor. Sandra provided endless
suggestions and help for all aspects of my career as a graduate
student.

The rest of my committee deserves thanks as well: Markus Dickinson,
Michael Gasser, and Steven Franks all contributed valuable advice and
suggestions while I was working on my dissertation, as well as during
my earlier work on English.

Other dialect researchers contributed to this dissertation: Henrik
Rosenkvist graciously gave me access to his current version of
Swediasyn. That data forms the basis of my test; without it, I would
not have a dissertation. Therese Leinonen donated her
high-resolution outline maps of Sweden and Swedish-speaking
Finland. This made my results look much better by taking advantage of
her hard work.

I should also thank the members of the Parsing Reading Group, both for
putting up with my crazy ideas and often suggesting equally crazy ones
in return. They helped me solve a number of practical problems with
the dissertation. Stephanie Dickinson was also a great help with
statistical tests. I recommend the services of the IU Statistical
Consulting Center to every graduate student at IU.

I want to thank the members of the IU Aikido Club and the North
Central church for helping me keep my body and spirit healthy
even while I was busy exercising my mind.

Finally, the Indiana University motto is ``Lux et veritas'', Latin for
``Light and truth''. I humbly acknowledge that God is the source of
all light and truth, and I hope that this dissertation reflects
a tiny part of that truth.



%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "dissertation.tex"
%%% End: 
