\documentclass{iuthesis}
% macros that I like go here.
\usepackage{graphicx}
\usepackage[all]{xy}
\usepackage{qtree}
\usepackage{robbib}
\includeonly{background,questions,methods,results,discussion}
\title{Syntactic Distance for Dialectometry}
\author{Nathan C. Sanders}
\advisor{Sandra K\"ubler}
\secondreader{Markus Dickinson}
\thirdreader{Steven Franks}
\fourthreader{Michael Gasser}

\department{Linguistics}
\submitdate{10/05/01} % a hopeful guess
\copyrightyear{2010}
\begin{document}

\begin{dedication}
  It's a great day for freedom
\end{dedication}
\begin{acknowledgements}
  I always thought Heeringa's acknowledgements were
  pretty cool
\end{acknowledgements}
\begin{abstract}
  Syntactic distance measures for dialectometry; tested on
  Swedish. It's all right.
\end{abstract}

\frontmatter
\maketitle
\signaturepage
\copyrightpage
\makededication
\makeack
\makeabstract
\tableofcontents
\mainmatter

\include{background}

\include{questions}

\include{methods}

\include{results}

\include{discussion}

\backmatter

\begin{center}
\section*{Nathan C. Sanders}
\end{center}
% \begin{tabular}{c|c|c} % re-arrange and maybe put home page on there?
% Electronic Mail & Address & Telephone \\
% \hline
% ncsander@indiana.edu & 4780 E St. Rd. 45 & 812-606-9785 \\
% sanders\_n@yahoo.com & Bloomington, IN 47408  & \\
% \end{tabular}

\subsection*{QUALIFICATIONS}
\begin{description}
  \item[Dialect Classification/Dialectology] Extraction of
    linguistic, human-interpretable features
  \item[Machine learning] Statistical (e.g. unsupervised clustering) \\
    Symbolic (e.g. learning and learnability in Optimality theory)
  \item[Parsing] Adaptation of computer science parsers to learner
    natural language
  \item[Programming] Fluent in Python, Haskell, Java/C\#, and Scheme \\
    Experience in C++, Perl, Visual Basic, Javascript, F\#, and
    Common Lisp \\
    Experience in prototyping and web programming \\
    Three-time competitor, ACM International Collegiate Programming
    Contest
\end{description}
% See http://www.sandersn.com/research.html for papers
% and code in the above areas.
\subsection*{EDUCATION}
\begin{description}
\item [PhD student in Linguistics, minor Computer Science] Indiana
  University, 2010
\item [MA Computational Linguistics] Indiana University, 2006, GPA 3.98
\item[BA Computer Science] minors in French and Spanish,
		College of the Ozarks, summa cum laude May 2004
\end{description}
\subsection*{EXPERIENCE}
\begin{description}
\item[August 2009--November 2009]
  SDET Intern, Microsoft, Bing Search Infrastructure --- Wrote
  software to suggest tests to developers at checkin. Tested
  the distributed computation system forming the infrastructure of
  Bing.
\item[August 2005--May 2009]
  Research Assistant, IU School of Medicine --- Investigated linguistic
  aspects of cochlear implant user development. Developed novel distance
  measure and compared it to existing measures and human judgments.
\item[May 2005--August 2005]
  Application Developer, IU Archives of Traditional Music --- Wrote Java
  application to gather cataloging data for videos. Worked with librarians and
  ethnomusicologists to gather archival metadata requirements.
\item[August 2004--May 2005]
  Web Programmer, IU Overseas Studies --- Maintained database and its
  web interface. Developed new features on request and maintained
  office computers.
\item[Summer 2004, Summer 2003]
  Lead Programmer, Everyware Inc --- Implemented web framework
  for interactive site building. Rewrote existing plug-in
  architecture and extended existing interface.
\item[Summer 2002]
  Intern Programmer, SIL International --- Programmed import
  process for translation editor in C++, with a pre-processor in Python.
\item[Spring 2001--Spring 2004]
  Lab Assistant, College of the Ozarks Foreign Language Lab ---
  Tutored students in Spanish and French. Maintained lab
  computers and created tracking databases.
\end{description}
\subsection*{PUBLICATIONS}

% no idea why I have to indent *twice*
\indent\indent Sanders, N. C. 2009. Phonological distance measures. \emph{Journal
    of Quantitative Linguistics}, 43:96--114.

  Sanders, N. C. 2008. Cluster analysis of phonological distance
  measures of cochlear implant users. In \emph{Proceedings of the
    Tenth International Conference on Cochlear Implants and Other
    Implantable Auditory Technologies}, 113.

 Sanders, N. C. 2007. Measuring Syntactic Difference in British
  English. In \emph{Proceeding of the ACL 2007 Student Research
    Workshop}, 1--6, Prague, Czech Republic, June.

  Sanders, N. C. and Chin, S. B. 2006. Phonological distance measures for cochlear implant users. \emph{Wiener Medizinische Wochenschrift 156}, [Suppl 119] 8

 Sanders, N. C.  2004. Compiler error detection techniques applied to natural language processing. In \emph{Proceedings of the Thirty-fifth SIGCSE Technical Symposium on Computer Science Education}. Association for Computing Machinery, 515

\bibliographystyle{robbib}
\bibliography{central}
\end{document}
