\chapter{Questions}
% TODO: Rewrite and merge the following question/hypothesis paragraph pairs
% H1 - organization is all wrong still
The state of syntax measures in dialectometry described above leaves
several research questions unresolved. The most important for this
proposal is whether $R$ is a good measure of syntax
distance. Specifically, have the ambiguous results of previous
research been a shortcoming of $R$, differences between phonological
and syntactic corpora, or differences between phonological and
syntactic dialect boundaries?

To investigate this, I propose Hypothesis 1: the features found by
dialectologists will agree with the highly ranked features used by $R$
for classification. I will test Hypothesis 1 by comparing $R$'s
results to the syntactic dialectology literature on Swedish. In
addition, Hypothesis 1B states that the regions of Sweden accepted by
dialectology will be reproduced by $R$. For example, my
previous research on British English reproduced the well-known North
England-South England dialect regions. However, this research will eliminate the
corpus variability in that research \cite{sanders08b} that resulted in
the confounding factors mentioned above, meaning that more precise
results, such as specific identifying features, should be detectable as well.

Finally, if $R$ is found to be a bad measure of syntax distance, this
dissertation will propose and evaluate alternative syntactic distance
measures. Specifically, $R$ is one way to aggregate features that are
created by decomposing sentences. It treats features as atomic, and
does not manipulate them in any syntax-specific ways. As such, $R$ is
not much different than Goebl's WIV. This may not be a problem if the
decomposition methods used to generate features adequately capture
dialect differences in independent, atomic features. If dialect
differences cannot be captured by independent, atomic features, then a
more syntax-specific method of combining features will be needed
instead. Alternatively, a more complex statistical measure may be
useful, taking the basic idea of $R$ and increasing its
sensitivity. For example, Kullbeck-Leibler divergence, like $R$,
provides a dissimilarity that is intuitively similar to distance.

%H2 - Dad didn't understand that this is other features to be fed into
%R not replacement of R entire.
A secondary question, relevant once a useful syntax distance measure
is established, is what input features cause $R$ to produce the best
results.  Previous work has shown that leaf-ancestor paths provide a
small advantage over part-of-speech trigrams, presumably by capturing
syntactic structure higher in the parse tree. Additional possible
feature sets include variations on the previously investigated
trigrams and leaf-ancestor paths, along with various kinds of backoff,
for example, to bigrams or coarser node tags. Features from dependency
parses may be useful, too, in capturing non-local dependencies that
can be captured neither by trigrams nor leaf-ancestor paths.

Therefore, I propose Hypothesis 2: better input features
for $R$ will produce more accurate syntax
distances. These features can be discovered by comparing performance
of a number of different feature sets on a fixed corpus. In addition,
combinations of successful features will produce even better
performance.
% This sentence is either redundant or should appear earlier.
The quality of a set of features can be
measured by its sensitivity---the number of significant distances it
finds---and the similarity of the highly ranked features $R$ produces
to those found by dialectologists.

% H3
A third question is whether $R$ agrees with phonological distance
measures like Levenshtein distance. Unlike agreement with traditional
dialectology, there is no {\it a priori} reason to expect agreement
between phonology and syntax in delineating dialect
boundaries. However, agreement with phonological distance would be
further evidence for the suitability of $R$ as a syntactic distance
measure.

Therefore, I propose Hypothesis 3: a phonology corpus and syntax corpus
constructed from the same data will provide better correlation between
phonology and syntax distance measures than a phonology corpus and
syntax corpus drawn from different data sources. I will test this
hypothesis by comparing results to my previous work on British English
phonological and syntactic corpora; there, no significant correlation
was found between the regions extracted from the two corpora drawn
from different data. If significant correlation is found by using the
same set of data for both corpora, it indicates that phonology and
syntax boundaries do coincide but that the agreement is weak enough to
be lost when using corpora collected from different populations.

\section{Question 1 : Agreement with Dialectology}

The first question is whether $R$ agrees with dialectology. This
question can be asked in a couple of different ways. The first (and
most difficult) is whether the features agree with the features
discovered by (human) dialectologists.

The others are whether boundaries agree, whether regions agree and
whether distances agree. However, traditional dialectology doesn't
produce distances much, and regions are usually only for particularly
strong distinctions (and are often derived from a set of strong
boundaries), so boundaries will be the most important of these
three.

Features is the most important for two reasons. First, it explains the
most and is the easiest to disprove. Second, dialectology will
produce the most instances of this; the others require a lot of work
on abstraction from dialectologists and (frankly) there aren't a lot
of Swedish syntactic dialectologists out there. I think there's two.

\subsection{Features}

We are looking for things like ``Swedish apparent cleft produces
sentences like this''. Then convert the examples they give into the
crappy formalism we use (implicitly), rip the sentence up into
features and see if those features show up in the mentioned
regions. If they do, how important are they? What features are more
important than they are?

There's probably one or two other things I should say about this. The
above paragraph will probably expand to about three on its own,
though.

\subsection{Boundaries}

Boundaries are what cause regions to emerge, and they are the
principal output of dialectology's attempt to find isoglosses. So I
would expect to find a lot of boundaries that overlap badly and don't
really form good regions. In fact, there isn't a whole lot of
difference between isogloss bundles and isoglosses for the purposes of
this dissertation; they both compare as directly to the features I
extract.

So, uh, this is really just a different way to say the same thing as
the Features section. I guess I should just stop talking now. (This
section should probably go above regions.)

\subsection{Regions}

Regions is the most straightforward type of matching: just use MDS or
clustering to get my results to show some kind of boundary, then check
dialectology for a list of regions. See if and how well they match.

The only complication is partial matches and weak boundaries;
statistical boundaries necessarily vary in strength, so the cutoff for
humans deciding there is a region here may differ from the cutoff they
use when looking at statistical results. And of course partial matches
just mean that it's less obvious how to evaluate objectively the
quality of the match. Perhaps you can just use $km^2$ overlap. Cheesy
but effective!

Matching specific regions is important. For instance, J\"amtland
appears to be reproduced by early results. Unfortunately, as is, there
don't seem to be many other SUPER STRONG boundaries mentioned with
things that aren't sister languages to Swedish instead of dialects
(technically, although you could easily analyse them the same way if
you want, but probably nobody is interviewing that way).

\subsection{Distances}

Since dialectology is qualitative, this necessarily is extremely
rough; we are looking for statements like ``there are numerous
differences between dialect X and Standard Swedish'' or ``dialect X is
very distant (or nearly indistinguishable) from Standard Swedish''.

\subsection{Alternate Distance Measures}

$R$ is the simplest measure. But there are others. However, R appears to
be the best, at least most consistently, over the largest variety of
features. In any case, $R$ is the more important question since a
distance measure is harder to change than a feature set. This isn't
true at the code level, but at the conceptual level, feature
sets are a lot easier to specify (copy from the proposal the 10
examples I came up with in 5 minutes).

So. Alternate distance measures. There are two directions worth
exploring: more complicated statistical measures, and more syntax
specific measures. There are more of the first and they fit more
naturally with the available data. But syntax specific measures would
be pretty cool. The problem is that most statistical syntax specific
measures can also be modelled as different feature sets, while keeping
the measure the same (more on features below).

Complicated measures are all around. WIV is one, although not really
designed for large data sets. But there are many others, like relative
entropy (Kullbeck-Leibler divergence?) and its variants
(Jenson-Shannon divergence). Probably you get could something out of
mutual information or something. I'd have to check the equation and
some commentary on it. Basically, any statistical measure of
divergence is suitable, and there don't seem to be a shortage of
those. I bet I could check Wikipedia to get a giant list that some cog
sci guys sat down and made.

\subsubsection{Syntax-specific distance measures}

Syntax-specific measures are harder. They have to somehow incorporate
syntax in the measure itself, in a way that can't be extracted and
reified as features for a generic measure. But this is hard for
statistical measures because syntax doesn't have much specific about
it; you have input divided into sentences, and each sentence having an
associated tree, and that's about it. Unlike phonology, you don't get
aligned words that you could use to compare the structures (trees)
directly. You don't get a well-specified several-layer model like you
do from Chomsky-Halle/Kiparsky-Archangeli.

Working by analogy from phonology, word order is equivalent to segment
order. If you want more precision than words (or better, trigrams),
model leaf-ancestor paths (and others) as bags of features for each
word. Then define a distance measure between these bags of features.

The problem with this is that, unless sentences are aligned (which
they aren't), then you'll have to compare all sentences to all
sentences. In that case any real differences will be lost in the noise
of high-distance sentences for which zero or one words match. Maybe
you could only look at sentences whose words match more than some
threshold and then compare the syntax features for each word.

\subsubsection{Important note on terms}

Note: divergence is a dissimilarity that is not necessarily
symmetric. Dissimilarity is a distance that is not necessarily
following the triangle inequality. I think, I should probably make
sure I'm not missing a property. Here, a ``distance measure'' really
only needs to be a dissimilarity.

\section{Question 2: Best Features}

Since the total question of this dissertation is whether syntax
distance can be made to work for dialectology, we must remember that
this is a two part question. The most important part is whether a
distance measure like $R$ can be found that works with arbitrary
features extracted from a large unaligned corpus. Once this is
established (dealt with in the previous section), the question shifts to what are
the best features.

This is an easier question, because feature sets are easier to dream
up. They are easier to combine (although I haven't done so yet.)
They are easier to tweak slightly for a great improvement in
results. The real problem with variation of feature sets is that there
no standard from syntax that can be easily adapted to
dialectometry. Phonology has order (mentioned above) and distinctive
features.

Previous work has shown that leaf-ancestor paths own a small advantage
over trigrams, and dependency paths seem to own advantages for
different measures than leaf-ancestor paths. So this question breaks
down into two pieces. How do you propose new feature sets, and how do
you tell the good ones?

\subsection{Proposing Feature Sets}

New feature sets are pretty easy to propose. Basically, chop up the
tree or dependency graph in some way, or condense the information down
in some way.  Trigrams take only the leaves, three at a
time. Leaf-ancestor paths take vertical slices of the tree. Dependency
paths take ``vertical'' slices of the graph (either arcs or nodes;
though arcs don't really work).

More lossy ways of dividing the parse might also be useful;
convolution kernels sum the differences between two trees. Oh wait, so
that wouldn't work. But something similar might. Maybe. Besides
convolution kernels, there are methods that try to identify the most
important features of the sentence manually. For example, the first or
last $n$ words in the sentence, words surrounding the predicate. Or
even lossier things like sentence length or tree height or number of
internal nodes, which captures the relative syntactic complexity of a
sentence.

Each of these has its advantages and disadvantages; leaf-ancestor
paths are good at capturing upper (hidden) structure. On the other
hand, dependency paths capture similar information but with more
emphasis on long-distance relations between words. Convolution
kernel-esque thing that I define tries to sum up the whole sentence in
a single symbol/number, as do tree height and node count. Other
specific features like first $n$ words intend to capture some aspect
of processing that will be reflected in real differences between
dialects.

\subsection{Evaluating Feature Sets}

The ideal feature set provides distances that match closely to the
real dialect differences between regions. Since these are not
precisely know, a number of proxies are necessary. One important proxy
is geographical distance; although there is no {\it a priori} reason
for geographical distance to match dialect distance, the dialectology
literature overwhelmingly shows that it does. Other proxies are
specific boundary features (mentioned above, for evaluating distance
measure). Also phonological distance.

Also, statistical significance. This is important but doesn't tell you
if you have a {\it good} set of features, only if you have a {\it bad}
set.

\subsection{Total Quality}

Total quality of the syntax distance is the multiple of distance measure
and feature set. Current results show inconsistency between measures
and sets; some combinations of measure and feature set work well, but
changing one makes it perform badly. In other words, one feature set
does not perform overwhelmingly better than the others; each feature
set has at least one measure for which it performs badly. (The
current patterns are odd.)


%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "dissertation.tex"
%%% End: 
